Modeling the impact of IT- and AI-related emissions in a university setting

### 2025-08-02 doc/paper.tex
Checkpoint
Intro

\addbibresource{ref.bib}
\title{Modeling the impact of IT- and AI-related emissions in a university setting.}
\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}

@@

\section{Abstract}
Information technology (IT) plays a major role in society and has non-negligible environmental impact, accounting for over 5\% of global electricity consumption and about 2\% of global carbon emissions. At the same time, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates. Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030. Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap. As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and analyze the impact of IT and AI on the university's carbon footprint. In this paper we present a simple interactive model of IT emissions on a univeristy campus. The model approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs. Our analysis shows that the embodied emissions of IT infrastructure and personal devices are a major component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.

@@

approximately $6.3\%$ of total emissions in 2024 \cite{uzh:sustainability:report},

artificial intelligence (AI) applications are being adopted at historically high rates exceeding those of the adoption of the personal computer and the internet \cite{bick:ai:adoption}.

We consider only direct impacts of AI: the increased electricity consumption and resulting emissions. We omit indirect impacts of AI adoption, such as potential emissions reductions due to improved building management: at the moment they are too nebulous to predict.

We attempt to model full lifecycle impact, including manufacture, operation, and disposal. We include not only university-owned devices (servers, classroom PCs, etc.), but also student and staff personal devices, since they represent the bulk of on-campus devices.

Likewise, we attempt to model full life cycle impact for cloud infrastructure and for LLMs.

We do not consider potential benefits of AI.

\section{Modeling methodology}

@@

For LLMs, we estimate operational energy use (Wh per query) using recent public data. Non-operational emissions, including embodied emissions of the server fleet and the cost of training the models, are harder to estimate and attribute. Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.
\subsection{Parameters and assumptions}

\section{Model details}

@@

\section{Results}

@@

\section{Recommendations}

@@

\section{Related work}

>>>>

\addbibresource{ref.bib}
\title{Modeling the impact of IT- and AI-related emissions in a university setting}
\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}

@@

\section{Abstract}
Information technology (IT) plays a major role in society and has non-negligible environmental impact, accounting for over 5\% of global electricity consumption and about 2\% of global carbon emissions. Moreover, artificial intelligence (AI) applications, and especially large language models (LLMs), are being adopted at historically high rates. Their growth and relatively high energy requirements are expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030. Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap. As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and analyze the impact of IT and AI on the university's carbon footprint. We present a simple interactive model of IT emissions on a univeristy campus. The model approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs. Our analysis shows that the embodied emissions of IT infrastructure and personal devices are a major component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.

@@

Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}. At the University of Zürich, which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt CO$_2$e of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}. The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).

AI applications are being adopted at historically high rates, more quickly than the personal computer or the internet were adopted in the past~\cite{bick:ai:adoption}. Enterprise AI adoption has accelerated, with generative AI tools seeing widespread deployment across industries within months of their release~\cite{mckinsey2024ai}. In July 2025, ChatGPT received 2.5 billion queries per day, up from 1 billion in December 2024, and within an order of magnitude of Google's approximately 15 billion queries per day~\cite{techcrunch:chatgpt}. AI growth is currently projected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030~\cite{iea:ai:energy}.

The University of Zürich has set itself the target of becoming climate-neutral by 2030 within the categories of energy, transportation, food, and waste, with additional goals in 2035 and beyond~\cite{uzh:sustainability:report}. Given the current sizable contribution of IT to the university's emissions, and the potential for growth through AI adoption in the coming years, we felt it was necessary to understand IT emissions and their growth trajectories in more detail.

This paper presents a simple interactive model of IT emissions on a university campus. We developed it in the specific context of the University of Zürich, but believe that it is parameterizable and general enough to apply to other similar university campuses or large corporate facilities.

We seek to answer three broad questions:
\begin{itemize}
  \item What are the main contributors to the IT emissions of a university?
  \item Could migrating some on-campus IT services to commercial cloud providers decrease the overall campus footprint?
  \item What is the likely impact of the surge in AI adoption? In particular, will the emissions impact of AI be so large as to make other sources of IT emissions negligible?
\end{itemize}

Our model sets relatively broad system boundaries. It takes into account university-owned equipment but also university services hosted by commercial cloud providers and university use of commercial generative AI products. We have also parameterized it to account for student devices not owned by the university, since they are key to student work and constitute the bulk of the university's end-user devices. We attempt to model devices' full lifecycle impact, including manufacture and disposal.

Experts rightly point to the many potential benefits of AI in the areas of climate change mitigation and adapation~\cite{climate:change:ai}. On a university campus, these might include more efficient building heating and cooling, improved transportation logistics, etc. However, those benefits are still to come and difficult to predict. In this paper, we consider only the direct impacts of AI in terms of increased electricity consumption and the resulting emissions.

The rest of this paper is organized as follows. Section~\ref{sec:model:methodology} presents the general modeling framework, including system boundaries and assumptions. Section~\ref{sec:model:details} justifies some of the specific parameter values used in the model. Section~\ref{sec:results} describes results and explores different scenarios and assumptions. Section~\ref{sec:recommendations} offers recommendations to university IT administrators. Section~\ref{sec:related:work} describes related work in this area.

\section{Modeling methodology}
\label{sec:model:methodology}

@@

For LLMs, we estimate operational energy use (Wh per query) using recent public data. Non-operational emissions, including embodied emissions of the server fleet and the cost of training the models, are harder to estimate and attribute. Based on data available for ChatGPT, we assume that non-operational emissions are a $40\%$ overhead on top of operational emissions.

\subsection{Parameters and assumptions}

\section{Model details}
\label{sec:model:details}

@@

\section{Results}
\label{sec:results}

@@

\section{Recommendations}
\label{sec:recommendations}

@@

\section{Related work}
\label{sec:related:work}
